thesis2024

### 2025-04-09 appendix1.qmd
Daily work session. SEM-EDS graph code and main paper graphs. SOM graphs start.

library(readr)
library(dplyr)
library(ggplot2)
library(janitor)

@@

>>>>

library(readr)
library(dplyr)
library(ggplot2)
library(janitor)
library(tidyr)
library(ggpmisc)

@@

```{r}
#| label: sem-graphs-som

# Prepare the energy ranges
energy_ranges <- list(
  C = c(0.2, 0.3),
  O = c(0.5, 0.6),
  Al = c(1.48, 1.5),
  Si = c(1.7, 1.8),
  P = c(2, 2.1),
  S = c(2.3,),
  Ca = c(3.68, 3.7),
  Ca_beta = c(4.01, 4.02),
  Ti = c(4.5, 4.6),
  Ti_beta = c(4.9, 5),
  Ti_L = c("0.45", "0.47"),
  Fe = c(6.38, 6.41),
  Fe_beta = c(7.05, 7.07),
  Fe_L = c(0.71, 0.73))

# M2_3_map2

M2_3_map2 <- read_csv('data/sem_data/M2_3_map2.csv')
filtered_M2_3_map2 <- subset(M2_3_map2, Energy >= 0.2 & Energy < 9)

# Extract the exact peak energy and counts to place labels
peaks_list_M2_3_map2 <- list()
for (element in names(energy_ranges)) {
  range_M2_3_map2 <- energy_ranges[[element]]
  subset_M2_3_map2 <- subset(filtered_M2_3_map2, Energy >= range_M2_3_map2[1] & Energy <= range_M2_3_map2[2])
  if (nrow(subset_M2_3_map2) > 0) {
    max_row_M2_3_map2 <- subset_M2_3_map2[which.max(subset_M2_3_map2$Counts), ]
    peaks_list_M2_3_map2[[element]] <- data.frame(
      Element = element,
      Energy = max_row_M2_3_map2$Energy,
      Counts = max_row_M2_3_map2$Counts
    )
  }
}

peaks_M2_3_map2 <- do.call(rbind, peaks_list_M2_3_map2)
rownames(peaks_M2_3_map2) <- NULL
peaks_M2_3_map2 <- peaks_M2_3_map2 |> 
  filter(Counts > 100)

M2_3_map2_graph <- filtered_M2_3_map2 |> 
  select(Energy, Counts) |>
  ggplot(aes(x = Energy, y = Counts)) +
  geom_area(fill = "snow3", alpha = 0.7) +
  geom_line(linewidth = 0.5, alpha = 0.5, linetype = "solid", colour = "#8B8989") +
  labs(x = "Energy (keV)", y = "Intensity (Counts)") +
  theme_minimal() +
  geom_point(data = peaks_M2_3_map2, aes(x = Energy, y = Counts, colour = Element), size = 3, shape = 18) +
  geom_label(data = peaks_M2_3_map2, aes(x = Energy, y = Counts, label = Element, colour = Element), size = 3, fill = "white", label.size = 0.2, vjust = -0.5, show.legend = FALSE) +
  theme(legend.position="none")

# M3_1

M3_1 <- read_csv('data/sem_data/M3_1_map1_map2.csv')
filtered_M3_1 <- subset(M3_1, Energy >= 0.2 & Energy < 9)

# Extract the exact peak energy and counts to place labels
peaks_list_M3_1 <- list()
for (element in names(energy_ranges)) {
  range_M3_1 <- energy_ranges[[element]]
  subset_M3_1 <- subset(filtered_M3_1, Energy >= range_M3_1[1] & Energy <= range_M3_1[2])
  if (nrow(subset_M3_1) > 0) {
    max_row_M3_1 <- subset_M3_1[which.max(subset_M3_1$Counts), ]
    peaks_list_M3_1[[element]] <- data.frame(
      Element = element,
      Energy = max_row_M3_1$Energy,
      Counts = max_row_M3_1$Counts
    )
  }
}

peaks_M3_1 <- do.call(rbind, peaks_list_M3_1)
rownames(peaks_M3_1) <- NULL
peaks_M3_1 <- peaks_M3_1 |> 
  filter(Counts > 100)

M3_1_graph <- filtered_M3_1 |> 
  select(Energy, Counts) |>
  ggplot(aes(x = Energy, y = Counts)) +
  geom_area(fill = "snow3", alpha = 0.7) +
  geom_line(linewidth = 0.5, alpha = 0.5, linetype = "solid", colour = "#8B8989") +
  labs(x = "Energy (keV)", y = "Intensity (Counts)") +
  theme_minimal() +
  geom_point(data = peaks_M3_1, aes(x = Energy, y = Counts, colour = Element), size = 3, shape = 18) +
  geom_label(data = peaks_M3_1, aes(x = Energy, y = Counts, label = Element, colour = Element), size = 3, fill = "white", label.size = 0.2, vjust = -0.5, show.legend = FALSE) +
  theme(legend.position="none")

### 2025-04-09 geochemistry.qmd
Daily work session. SEM-EDS graph code and main paper graphs. SOM graphs start.

library(readr)
library(dplyr)
library(janitor)
library(tidyr)
library(gt)
library(ggplot2)
library(cowplot)
library(tidyverse)
library(ggfortify)
library(FactoMineR)
library(factoextra)
library(ggpmisc)

@@

@@

knitr::include_graphics(here::here("figures/within-chert/M18_7_M18_8.png"))

>>>>

library(readr)
library(dplyr)
library(janitor)
library(tidyr)
library(gt)
library(ggplot2)
library(cowplot)
library(tidyverse)
library(ggfortify)
library(FactoMineR)
library(factoextra)
library(ggpmisc)
library(magick)

@@

```{r}
#| label: sem-graphs

# Prepare the energy ranges
energy_ranges <- list(
  C = c(0.2, 0.3),
  O = c(0.5, 0.6),
  Al = c(1.48, 1.5),
  Si = c(1.7, 1.8),
  P = c(2, 2.1),
  Ca = c(3.68, 3.7),
  Ca_beta = c(4.01, 4.02),
  Ti = c(4.5, 4.6),
  Ti_beta = c(4.9, 5),
  Ti_L = c("0.45", "0.47"),
  Fe = c(6.38, 6.41),
  Fe_beta = c(7.05, 7.07),
  Fe_L = c(0.71, 0.73))

# M18_8

M18_8 <- read_csv('data/sem_data/M18_8.csv')
filtered_M18_8 <- subset(M18_8, Energy >= 0.2 & Energy < 9)

# Extract the exact peak energy and counts to place labels
peaks_list_M18_8 <- list()
for (element in names(energy_ranges)) {
  range_M18_8 <- energy_ranges[[element]]
  subset_M18_8 <- subset(filtered_M18_8, Energy >= range_M18_8[1] & Energy <= range_M18_8[2])
  if (nrow(subset_M18_8) > 0) {
    max_row_M18_8 <- subset_M18_8[which.max(subset_M18_8$Counts), ]
    peaks_list_M18_8[[element]] <- data.frame(
      Element = element,
      Energy = max_row_M18_8$Energy,
      Counts = max_row_M18_8$Counts
    )
  }
}

peaks_M18_8 <- do.call(rbind, peaks_list_M18_8)
rownames(peaks_M18_8) <- NULL
peaks_M18_8 <- peaks_M18_8 |> 
  filter(Counts > 100)

M18_8_graph <- filtered_M18_8 |> 
  select(Energy, Counts) |>
  ggplot(aes(x = Energy, y = Counts)) +
  geom_area(fill = "snow3", alpha = 0.7) +
  geom_line(linewidth = 0.5, alpha = 0.5, linetype = "solid", colour = "#8B8989") +
  labs(x = "Energy (keV)", y = "Intensity (Counts)") +
  theme_minimal() +
  geom_point(data = peaks_M18_8, aes(x = Energy, y = Counts, colour = Element), size = 3, shape = 18) +
  geom_label(data = peaks_M18_8, aes(x = Energy, y = Counts, label = Element, colour = Element), size = 3, fill = "white", label.size = 0.2, vjust = -0.5, show.legend = FALSE) +
  theme(legend.position="none")

# M18_7

M18_7 <- read_csv('data/sem_data/M18_7.csv')
filtered_M18_7 <- subset(M18_7, Energy >= 0.2 & Energy < 9)

# Extract the exact peak energy and counts to place labels
peaks_list_M18_7 <- list()
for (element in names(energy_ranges)) {
  range_M18_7 <- energy_ranges[[element]]
  subset_M18_7 <- subset(filtered_M18_7, Energy >= range_M18_7[1] & Energy <= range_M18_7[2])
  if (nrow(subset_M18_7) > 0) {
    max_row_M18_7 <- subset_M18_7[which.max(subset_M18_7$Counts), ]
    peaks_list_M18_7[[element]] <- data.frame(
      Element = element,
      Energy = max_row_M18_7$Energy,
      Counts = max_row_M18_7$Counts
    )
  }
}

peaks_M18_7 <- do.call(rbind, peaks_list_M18_7)
rownames(peaks_M18_7) <- NULL
peaks_M18_7 <- peaks_M18_7 |> 
  filter(Counts > 100)

M18_7_graph <- filtered_M18_7 |> 
  select(Energy, Counts) |>
  ggplot(aes(x = Energy, y = Counts)) +
  geom_area(fill = "snow3", alpha = 0.7) +
  geom_line(linewidth = 0.5, alpha = 0.5, linetype = "solid", colour = "#8B8989") +
  labs(x = "Energy (keV)", y = "Intensity (Counts)") +
  theme_minimal() +
  geom_point(data = peaks_M18_7, aes(x = Energy, y = Counts, colour = Element), size = 3, shape = 18) +
  geom_label(data = peaks_M18_7, aes(x = Energy, y = Counts, label = Element, colour = Element), size = 3, fill = "white", label.size = 0.2, vjust = -0.5, show.legend = FALSE) +
  theme(legend.position="none")

# M4_3

M4_3 <- read_csv('data/sem_data/M4_3.csv')
filtered_M4_3 <- subset(M4_3, Energy >= 0.2 & Energy < 9)

# Extract the exact peak energy and counts to place labels
peaks_list_M4_3 <- list()
for (element in names(energy_ranges)) {
  range_M4_3 <- energy_ranges[[element]]
  subset_M4_3 <- subset(filtered_M4_3, Energy >= range_M4_3[1] & Energy <= range_M4_3[2])
  if (nrow(subset_M4_3) > 0) {
    max_row_M4_3 <- subset_M4_3[which.max(subset_M4_3$Counts), ]
    peaks_list_M4_3[[element]] <- data.frame(
      Element = element,
      Energy = max_row_M4_3$Energy,
      Counts = max_row_M4_3$Counts
    )
  }
}

peaks_M4_3 <- do.call(rbind, peaks_list_M4_3)
rownames(peaks_M4_3) <- NULL
peaks_M4_3 <- peaks_M4_3 |> 
  filter(Counts > 100)

M4_3_graph <- filtered_M4_3 |> 
  select(Energy, Counts) |>
  ggplot(aes(x = Energy, y = Counts)) +
  geom_area(fill = "snow3", alpha = 0.7) +
  geom_line(linewidth = 0.5, alpha = 0.5, linetype = "solid", colour = "#8B8989") +
  labs(x = "Energy (keV)", y = "Intensity (Counts)") +
  theme_minimal() +
  geom_point(data = peaks_M4_3, aes(x = Energy, y = Counts, colour = Element), size = 3, shape = 18) +
  geom_label(data = peaks_M4_3, aes(x = Energy, y = Counts, label = Element, colour = Element), size = 3, fill = "white", label.size = 0.2, vjust = -0.5, show.legend = FALSE) +
  theme(legend.position="none")

# C: 0.2-0.3
# O: 0.5-0.6
# Al: 1.48-1.5
# Si: 1.7-1.75
# P: 2-2.1
# Ti: 4.5-4.6 and 4.9-5

  #coord_cartesian(ylim = c(-5000, 55000))+
  #geom_point(data = sp15_marks, aes(x = Angle, y = Intensity + 600, colour = Phase), size = 3, shape = 18) +
  #geom_label(data = sp15_marks, aes(x = Angle, y = Intensity, label = d), 
    #        size = 3, fill = "white", label.size = 0.2, vjust = -1, show.legend = FALSE) +
  #scale_colour_manual(values = c("lightsteelblue3", "salmon", "steelblue1" ,"orchid","snow4")) +
  #annotate(geom = "table", x = 75, y = 55000, label = list(cbind(sp15_perc)))

@@

#knitr::include_graphics(here::here("figures/within-chert/M18_7_M18_8.png"))
M18_7_M18_8_fig <- magick::image_read(here::here("figures/within-chert/M18_7_M18_8.png"))
M18_7_M18_8_gg <- cowplot::ggdraw() + cowplot::draw_image(M18_7_M18_8_fig)

M18_7_M18_8_plots <- plot_grid(M18_7_graph, M18_8_graph, labels = c(1, 2))

plot_grid(M18_7_M18_8_gg, M18_7_M18_8_plots, ncol = 1, rel_heights = c(1, 1))

### 2025-04-09 thesis2024.tex
Daily work session. SEM-EDS graph code and main paper graphs. SOM graphs start.

\chapter{geochemistry}\label{geochemistry}

@@

The pXRF measured several major and minor elements, of which a small
amount returned values of 0 or were below the limit of detection
(\textless LOD). These were uranium (U), thorium (Th), bismuth (Bi),
thallium (Tl), mercury (Hg), platinum (Pt), tantalum (Ta), hafnium (Hf),
lanthanum (La), antimony (Sb), Tin (Sn), cadmium (Cd), rhodium (Rh),
niobium (Nb), selenium (Se), zinc (Zn), nickel (Ni), cobalt (Co) and
chromium (Cr). They were removed from the analysis based on their
nonexistence, although they can still be found in the raw pXRF results.

@@

Some supplementary materials.

>>>>

\chapter{Within chert: a multi-technique mineral and geochemical
approach to the study of chert of southwestern
Iberia}\label{within-chert-a-multi-technique-mineral-and-geochemical-approach-to-the-study-of-chert-of-southwestern-iberia}

@@

In this section we present the summarised results, focusing on the main
observed patterns and differences between formations and archaeological
chert types. A detailed description of the SEM-EDS results for each
formation and chert type and table of occurrences for identified
elements and minerals can be found in the SOM (++).

All analysed samples are mainly composed of quartz; in all samples, the
atomic percentage of oxygen is approximately double of the atomic
percentage of silica, showing a 2:1 ratio (table +).

Similarly, the presence of skeletal remains, represented by peaks of
calcium, phosphorus and frequently accompanied by small peaks of
sulphur, shows no patterns and occurs in samples from all formations and
chert types. Although frequently without a discernible shape, oval
fossils (although currently unidentified) were observed in LJW
geological samples and archaeological samples from local cherts (T2 and
T4). This type of large, oval fossil, also composed of calcium,
phosphorus and sulphur, seems to be limited to local cherts and to
specific areas of nodules, characterised by high amounts of these
skeletal remains.

\includegraphics{figures/within-chert/M3_3_map1.png}

\caption{\label{fig-skeletal}Skeletal remain from sample SP16\_BLX, from
the Lower Jurassic ourcrops of western Algarve. A: Detail of a
phosphatized skeletal remain composed by calcium (A1), phosphorus (A2)
and a small peak of sulphur (A3). The skeletal remain seems to be
oxidized, with small peaks of iron within (A4).}

\includegraphics{figures/within-chert/M15_4.png}

\caption{\label{fig-oval-fossils}A-C: General view and details from an
archaeological sample from local chert type (T4). A: general view of a
concentration of oval fossils. B: general view of a concentration of
oval fossils and other inclusions. C: detail of an irregular inclusion
composed of titanium (C1) and an oval phosphatised skeletal remain
composed mainly of calcium (C2), phosphorus (C3) and a small peak of
sulphur.}

By observing the surfaces of samples and the presence of skeletal
remains, some differences could be noted between local cherts and
non-local archaeological samples. In comparison to the frequent to
common skeletal remains in the Jurassic cherts from the Algarve, and
archaeological cherts attributed to local sources (T2-5), cherts from
non-local sources such as T6-8 show little porosity and rare presence of
inclusions. Whenever present, skeletal remains are small and show small
peaks (\textless3 cpv/ev).

\includegraphics{geochemistry_files/figure-pdf/fig-t6-general-1.pdf}

\caption{\label{fig-t6-general}A-D: General view and details from an
archaeological sample from non-local chert type (T6). A: general view of
the sample with uncommon inclusions. B: detail of an inclusion seen in
A, composed mostly of iron and small peaks of calcium and phosphorus,
making this inclusion a possible oxidised skeletal remains of a fossil.
C: general view of the sample with no visible inclusions. D: detail of a
small and irregular iron inclusion.}

The rarity of inclusions is not limited to skeletal remains, but to
other inclusions such as oxides, which in T6-8 are small and rare.
Despite this difference in abundance and size, oxides are mainly
composed of iron, often associated with other elements, possibly
suggesting the existence of iron oxide (Fe2O3), pyrite (FeS2) and
ilmenite (FeTiO3). Iron oxides were found both as inclusions scattered
in the samples, but also frequently associated with skeletal remains,
which may be related to their oxidization. Pyrite was also identified in
the shape of spherical pyrite framboids, as well as dispersed pyrite
nanocrystals. These are characterised not only by large peaks of iron
and associated sulphur, but also of round/angular small nanocrystals
organised in spherical shapes, semi-spherical or simply dispersed
crystals, frequently in association with spherical framboids. These were
present mainly in geological samples from the Upper Jurassic (both west
and eastern formations), but barely identified in archaeological
samples.

\includegraphics{figures/within-chert/M10_5_map1_map2.png}

\caption{\label{fig-framboids}A-B: Detail from the geological sample
SP59\_JOR from the Upper Jurassic outcrops (Jordana formation) of
eastern Algarve. A: detailed view of a spherical pyrite framboid
composed of iron (A1) and sulphur (A2). B: detailed view of an inclusion
composed of iron (B1) and sulphur (B2).}

In a single sample from Lower Jurassic and samples from local T4 (both
characterised by the previously mentioned oval fossils) inclusions
composed of titanium (without associated iron) were identified, which
may be related to the presence of rutile (TiO2). Small peaks of titanium
(\textless3 cpv/ev) were also identified in 3 out of 4 non-local T6
samples.

Other elements were identified through EDS maps and spectra (e.g.,
chlorine, potassium or magnesium), although their distribution also
showed no patterns between formations and chert types. The exception are
calcium and aluminium.

Whenever isolated, calcium can be found in high peaks in cherts from
Upper Jurassic formations and is probably related to the presence of
calcite inclusions.

\includegraphics{figures/within-chert/M4_3.png}

\caption{\label{fig-calcite}A-B: General view of calcite inclusions
(black arrows) from the geological sample SP69\_MAR from the Upper
Jurassic outcrops of western Algarve.}

Aluminium may be related to the speed of sedimentation; high peaks of
aluminum may be related to fastest sedimentations speeds and may be a
useful component to identify differences between cherts. Almost all
chert samples show aluminum peaks, although mostly small (\textless{} 3
cpv/ev). An exception are cherts from the Eastern basin of the Algarve
(MJE and UJE), samples from non-local archaeological T7 and T8. The
exception to the presence of aluminum is the chalcedony (both geological
from TW and archaeological group T1) and all archaeological samples from
the non-local group T6, with sources previously attributed to the
Cretaceous cherts from Central Portugal.

@@

\section{XRD results}\label{xrd-results}

In this subsection we present the individual spectra graphs for all XRD
samples, organised by geological formation, excluding those presented in
the main paper and the results analysed and interpreted at the
Geociencias Barcelona-CSIC. The original data necessary to produce these
visualisations can be found and freely accessed in our online
compendium, in the associated GitHub project under the path
``../data/xrd\_data''.

\includegraphics{appendix1_files/figure-pdf/fig-xrd-sp6-1.pdf}

\caption{\label{fig-xrd-sp6}Plot of minerals identified through XRD in
the Lower Jurassic (Carixian) chert sample from the outcrop Cabo S.
Vicente (SP6\_CSV). The rhombus shape signifies main peaks key to
identify each mineral. Labels represent the d spacing values of key
mineral peaks.}

\includegraphics{appendix1_files/figure-pdf/fig-xrd-sp7-1.pdf}

\caption{\label{fig-xrd-sp7}Plot of minerals identified through XRD in
the Lower Jurassic (Carixian) chert sample from the outcrop Cabo S.
Vicente (SP7\_CSV). The rhombus shape signifies main peaks key to
identify each mineral. Labels represent the d spacing values of key
mineral peaks.}

\includegraphics{appendix1_files/figure-pdf/fig-xrd-sp10-1.pdf}

\caption{\label{fig-xrd-sp10}Plot of minerals identified through XRD in
the Lower Jurassic (Carixian) chert sample from the outcrop Ponta dos
Altos (SP10\_PdA). The rhombus shape signifies main peaks key to
identify each mineral. Labels represent the d spacing values of key
mineral peaks.}

\includegraphics{appendix1_files/figure-pdf/fig-xrd-sp14-1.pdf}

\caption{\label{fig-xrd-sp14}Plot of minerals identified through XRD in
the Lower Jurassic (Carixian) chert sample from the outcrop Praia do
Belixe (SP14\_PBX). The rhombus shape signifies main peaks key to
identify each mineral. Labels represent the d spacing values of key
mineral peaks.}

\includegraphics{appendix1_files/figure-pdf/fig-xrd-sp18-1.pdf}

\caption{\label{fig-xrd-sp18}Plot of minerals identified through XRD in
the Lower Jurassic (Carixian) chert sample from the outcrop Praia do
Belixe (SP18\_PBX). The rhombus shape signifies main peaks key to
identify each mineral. Labels represent the d spacing values of key
mineral peaks.}

\includegraphics{appendix1_files/figure-pdf/fig-xrd-sp21-1.pdf}

\caption{\label{fig-xrd-sp21}Plot of minerals identified through XRD in
the Lower Jurassic (Carixian) chert sample from the outcrop Belixe Sul,
north of Praia do Belixe (SP21\_BLS). The rhombus shape signifies main
peaks key to identify each mineral. Labels represent the d spacing
values of key mineral peaks.}

\includegraphics{appendix1_files/figure-pdf/fig-xrd-sp28-1.pdf}

\caption{\label{fig-xrd-sp28}Plot of minerals identified through XRD in
the Lower Jurassic (Carixian) chert sample from the outcrop Aspa, north
of Cabo S. Vicente (SP28\_ASP). The rhombus shape signifies main peaks
key to identify each mineral. Labels represent the d spacing values of
key mineral peaks.}

\includegraphics{appendix1_files/figure-pdf/fig-xrd-sp30-1.pdf}

\caption{\label{fig-xrd-sp30}Plot of minerals identified through XRD in
the Lower Jurassic (Carixian) chert sample from the outcrop Foz dos
Fornos (SP30\_FZF). The rhombus shape signifies main peaks key to
identify each mineral. Labels represent the d spacing values of key
mineral peaks.}

\includegraphics{appendix1_files/figure-pdf/fig-xrd-sp33-1.pdf}

\caption{\label{fig-xrd-sp33}Plot of minerals identified through XRD in
the Lower Jurassic (Carixian) chert sample from the outcrop Foz dos
Fornos (SP33\_FZF). The rhombus shape signifies main peaks key to
identify each mineral. Labels represent the d spacing values of key
mineral peaks.}

\includegraphics{appendix1_files/figure-pdf/fig-xrd-sp34-1.pdf}

\caption{\label{fig-xrd-sp34}Plot of minerals identified through XRD in
the Lower Jurassic (Carixian) chert sample from the outcrop Ponta dos
Altos (SP34\_PdA\_a). The rhombus shape signifies main peaks key to
identify each mineral. Labels represent the d spacing values of key
mineral peaks.}

\includegraphics{appendix1_files/figure-pdf/fig-xrd-sp40-1.pdf}

\caption{\label{fig-xrd-sp40}Plot of minerals identified through XRD in
the Lower Jurassic (Carixian) chert sample from the outcrop Ferrel
(SP40\_b). The rhombus shape signifies main peaks key to identify each
mineral. Labels represent the d spacing values of key mineral peaks.}

\includegraphics{appendix1_files/figure-pdf/fig-xrd-sp50-1.pdf}

\caption{\label{fig-xrd-sp50}Plot of minerals identified through XRD in
the Middle Jurassic (Malhão formation) chert sample from the outcrop
Casal da Colina (SP50\_CdC). The rhombus shape signifies main peaks key
to identify each mineral. Labels represent the d spacing values of key
mineral peaks.}

\includegraphics{appendix1_files/figure-pdf/fig-xrd-sp52-1.pdf}

\caption{\label{fig-xrd-sp52}Plot of minerals identified through XRD in
the Middle Jurassic (Malhão formation) chert sample from the outcrop
Casal da Colina (SP52\_CdC). The rhombus shape signifies main peaks key
to identify each mineral. Labels represent the d spacing values of key
mineral peaks.}

\includegraphics{appendix1_files/figure-pdf/fig-xrd-sp57-1.pdf}

\caption{\label{fig-xrd-sp57}Plot of minerals identified through XRD in
the Upper Jurassic (Jordana formations) chert sample from the outcrop
Jordana (SP57\_JOR). The rhombus shape signifies main peaks key to
identify each mineral. Labels represent the d spacing values of key
mineral peaks.}

\includegraphics{appendix1_files/figure-pdf/fig-xrd-sp59-1.pdf}

\caption{\label{fig-xrd-sp59}Plot of minerals identified through XRD in
the Upper Jurassic (Jordana formation) chert sample from the outcrop
Jordana (SP59\_JOR). The rhombus shape signifies main peaks key to
identify each mineral. Labels represent the d spacing values of key
mineral peaks.}

\includegraphics{appendix1_files/figure-pdf/fig-xrd-sp61-1.pdf}

\caption{\label{fig-xrd-sp61}Plot of minerals identified through XRD in
the Upper Jurassic (Jordana formation) chert sample from the secondary
outcrop Peral (SP61\_PER). The rhombus shape signifies main peaks key to
identify each mineral. Labels represent the d spacing values of key
mineral peaks.}

\includegraphics{appendix1_files/figure-pdf/fig-xrd-sp63-1.pdf}

\caption{\label{fig-xrd-sp63}Plot of minerals identified through XRD in
the Middle Jurassic (Malhão formation) chert sample from the outcrop
Oliveiras (SP63\_OLV). The rhombus shape signifies main peaks key to
identify each mineral. Labels represent the d spacing values of key
mineral peaks.}

\includegraphics{appendix1_files/figure-pdf/fig-xrd-sp66-1.pdf}

\caption{\label{fig-xrd-sp66}Plot of minerals identified through XRD in
the Middle Jurassic (Malhão formation) chert sample from the outcrop
Malhão (SP66\_MALH). The rhombus shape signifies main peaks key to
identify each mineral. Labels represent the d spacing values of key
mineral peaks.}

\section{SEM-EDS results}\label{sem-eds-results}

\subsection{Geological samples}\label{geological-samples}

\subsubsection{Lower Jurassic of western
Algarve}\label{lower-jurassic-of-western-algarve}

The analysed samples from the Lower Jurassic outcrops are homogeneous
regarding their structure, inclusions and mineral compositions. The
analysis of their composition and inclusions shows no specific
differences between the outcrops, and they will be described as a single
group (Carixian, Lower Jurassic formations of Western Algarve, including
the inland outcrop of Ferrel).

They are mainly composed of cryptocrystalline quartz (predominantly
silica or SiO2), and small peaks of aluminium, with a single sample also
showing a high aluminium peak. Small peaks of chlorine and potassium can
also be identified in less than 50\% of the samples.

Phosphatised skeletal remains are very frequent, occurring in more than
80\% of the samples. These are all composed of calcium and phosphorus,
and frequently show an associated smaller peak of sulphur
(\textasciitilde50\% of the samples) or sulphur and fluorine
(\textasciitilde70\% of the samples). These fossils also frequently show
some oxidation, due to the associated presence of iron bands or
restricted areas within the phosphatised skeletal remains. These
skeletal remains are frequently irregular and indeterminate; however, a
small number of samples show phosphatised skeletal remains with
homogeneous oval shapes, which are currently indeterminate.

Finally, iron oxides are present in all the samples, with an iron (Fe)
composition. Other types of iron oxides are also present in lesser
frequencies, as equally large peaks of iron and titanium (40\% of the
samples), iron and fluoride (20\%), and titanium with a small peak of
iron (\textless10\%). A single sample shows a high titanium peak without
iron and another titanium with a smaller peak of phosphorus. Most of
these iron or titanium inclusions have irregular shapes and varying
sizes. Albeit rare (present in 3 samples), pyrite can be identified in
the form of spherical framboids or dispersed pyrite nanocrystals.
